\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-02-07 analysis/paper/paper.Rmd

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)
options(kableExtra.latex.load_packages = FALSE)
library(tidyverse)
library(magrittr)
library(survival)
library(casebase)
#devtools::load_all('../../../casebase')#testing
library(splines)
library(tibble)
library(glmnet)
library(data.table)
library(kableExtra)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(1234)

eval_introduction <- FALSE
eval_theory <- FALSE
eval_implementation <- FALSE
eval_cs1 <- FALSE
eval_cs2 <- FALSE
eval_cs3 <- TRUE
eval_cs4 <- FALSE
```

@@

In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base sampling *uniformly*, i.e. to sample the base series uniformly from the study base. In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base. More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.

The `casebase` package fits the family of hazard functions of the form

$$ h(t;X) = \exp[g(t;X)] $$
where $t$ denotes time and $X$, the individual's covariate profile. Different functions of $t$ lead to different parametric hazard models. The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.

$$ h(t;X) = \exp(\beta_0 + \beta_1 X) $$

The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the *memoryless property* (Kalbfleisch and Prentice, 2002).

@@

At the time of writing, a cursory examination of CRAN's task view on survival analysis reveals that there are over 250 packages related to survival analysis [-@survTaskView]. For the purposes of this article, we restricted our description to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and cumulative incidence curve estimation. By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages. These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}. In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models. The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.

Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche]. Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{CoxRidge} [-@perperoglou] and \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation for ridge. \pkg{casebase} provides penalized estimation in the context of parametric hazards. To our knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to offer this functionality. \textcolor{red}{I can dive deeper to determine if this is a fair statement}

Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@mahani2015bayesian], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival}. The types of models they allow vary for each package. For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of time using splines, which allows flexible modeling of the hazard function. Moreover, \pkg{flexsurv} has the ability to estimate both scale and shape parameters for a variety of parametric families (see Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear model of covariates (including time). Therefore, our package allows the user to model the effect of time using splines, and through interaction terms involving covariates and time, it also allows user to fit time-varying coefficient models. However, we do not explicitly model any shape parameter, unlike \pkg{flexsurv}.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, \pkg{cmprsk} and \pkg{survival} contain implementations for competing risks. The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression. Note that \pkg{CFC} has extra features outside the scope of the metthods discussed in this paper.

@@

\textbf{CFC} \cite{mahani2015bayesian}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
textbf{cmprsk} \cite{gray_2019}            & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\

@@

In the next case study, we show how case-base sampling can also be used in the context of a competing risk analysis. For illustrative purposes, we will use the same data that was used in Scrucca *et al* [-@scrucca2010regression]. The data was downloaded from the first author's website, and it is also available as part of the \pkg{casebase} package.

```{r bmtcrr-data, eval = eval_cs2, message = FALSE}

data(bmtcrr)
```

@@

```{r bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.", eval = eval_cs2}
library(ggplot2)
risk_all <- dplyr::bind_rows(
    data.frame(Time = time_points, 
               Method = "Case-base", 
               Risk = risk_cb[,2], 
               Disease = "ALL",
               stringsAsFactors = FALSE),
    data.frame(Time = time_points, 
               Method = "Case-base", 
               Risk = risk_cb[,3], 
               Disease = "AML",
               stringsAsFactors = FALSE),
    data.frame(Time = time_points,
               Method = "Fine-Gray",
               Risk = risk_fg$P1[1,],
               Disease = "ALL",
               stringsAsFactors = FALSE),
    data.frame(Time = time_points,
               Method = "Fine-Gray",
               Risk = risk_fg$P1[2,],
               Disease = "AML",
               stringsAsFactors = FALSE)
    )

@@

For the third case study, we will introduce regularization into the estimation of the hazard function. This can be performed using a combination of case-base sampling and regularized logistic regression. To demonstrate this capability, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support]. The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital. Imputation was conducted using the \pkg{mice} package in \proglang{R} with default settings [@mice]. Before imputation, there were a total of 1000 individuals and 35 variables. charges, Totcst, totmcst and sfdm2 were removed as they are response variables we are not interested in for the sake of this analysis. After imputation, we have 690 individuals. We lose individuals due to coliniarity between variables, preventing imputation. If a pair of covariates are deemed colinear to a certain degree, they are excluded from any imputations. Here, only edu  income  avtisst*  race  ph*  glucose*  adlp* were imputed. Of these 690 individuals, 66.09% experienced the event of interest. <!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}. The original data was obtained from the Department of Biostatistics at Vanderbilt University. The imputed dataset is available as part of the \pkg{casebase} package.

@@

<!--Hospital death as a covariate was removed, as this is directly informative of death. The glmnet package on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.--> As before, the first step is to visualize the incidence density. The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}.

```{r support-poptime, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data. Red points represent the case series. The gray space demonstrates the base series."}
data(support)
support_popTime <- popTime(support,
                           time = "d.time", event = "death",percentile_number = 0.31)
plot(support_popTime)
```

Before we proceed with the analysis, we first remove hospital death as a covariate, as this is directly informative of death. All covariates will be used to estimate the hazard function. The hazard function will be used to estimate the cumulative incidence of each individual in the study. Then, the average cumulative incidience curve for all observations will be used for our comparison.

```{r supportSetUp, echo = FALSE, eval = eval_cs3}
completeData=casebase::support
#remove hospital dead
completeData<-completeData[,-c(4)]
ratio=100  #universal ratio for eval_cs3
#keep one individual out, at random, for which we will develop an 
#absolute risk curve.
#sam=sample(1:nrow(completeData), 1)
#Set workingData to the remaining individuals in the data
#workingData=completeData[-c(sam),]
workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
#Create u and xCox, which will be used when fitting Cox with glmnet.
newData=workingCompleteData
x=workingCompleteData
#x and y will be used to fit the casebase model under glmnet.
#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y=data.matrix(completeData[,c(2,5)])
```

Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model. First, we fit a smooth hazard to the data using a weibull distribution. As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model. For the SUPPORT dataset, all factor variables were converted into indicator variables. Here, we used lasso penalization by setting alpha to 1.

```{r cbHazardInteractions}, echo = TRUE, eval = eval_cs3}

cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
                                       formula_time = ~log(d.time),alpha=1,ratio=ratio)
```
Then, we take our model and use the `absoluteRisk` function to integrate and retrieve our cumulative incidence curves.

```{r CBabsolute, echo = FALSE, eval = eval_cs3}
#Estimating the absolute risk curve using the newData parameter.
casebaseCI=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 20),newdata = x , s="lambda.1se",method=c("numerical"))

cbCIAve=data.frame(casebaseCI[,1],rowMeans(casebaseCI[,-c(1)]))
```
We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve. The cox regression absolute risk curve with lasso penalization required two extra steps to retrieve. Coxnet from the glmnet package has tools to fit a regularized cox model, but requires a survival object skeleton of the selected variables, so that the survival package tools can be used to retrive the absolute risk. We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit. This second model serves as a skeleton that permits the use of survival package functions. The coefficients from the regularized model will replace the coefficients in our skeleton. the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error. For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.

```{r coxHazAbsolute, echo = FALSE, eval = eval_cs3}
#Create u and xCox, which will be used when fitting Cox with glmnet.
u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))
#hazard for cox using glmnet
coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
#convergence demonstrated in plot
#plot(coxglmFit)
#taking the coefficient estimates for later use
nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
#creating a new dataset that only contains the covariates chosen through glmnet.
#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
#fitting a cox model using regular estimation, however we will not keep it.
#this is used more as an object place holder.
coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
#The coefficients of this object will be replaced with the estimates from coxglmFit.
#Doing so makes it so that everything is invalid aside from the coefficients.
#In this case, all we need to estimate the absolute risk is the coefficients.
#Std. error would be incorrect here, if we were to draw error bars.
coxFit$coefficients<-nonzero_coef_cox@x
#Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
newDataCox=newData[,nonzero_covariate_cox$X1-1]
abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(newDataCox),time = seq(0,max(completeData$d.time), 1),type="breslow")
abCF<-abCoxFit
abCF$surv<-rowMeans(abCF$surv)
```
We used the survival package to calculate cox regression's cumulative incidience without penalization and the kaplan-meier curve for incidience.

```{r unpenalizedCox, echo = FALSE, eval = eval_cs3}
Cox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,-c(1)]))
#coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
#abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
#abCFup<-abcoxFitup
#abCFup$surv<-rowMeans(abCFup$surv)
#cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
       #                                ratio=rat,data=fullDataCox)
#summary(cb.ModelNorm)#need to figure out how to deal with NAs
#casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
#cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))
```

```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}

coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)

#creating a surv object to be used to fit an unadjusted absolute risk curve.
# (Kaplan-Meier risk curve)
km <- survival::Surv(time = completeData$d.time, event = completeData$death)
abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
```

Now that our four cumulative incidence estimations have been calculated, we compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
abCFup<-abcoxFitup
abCFup$surv<-rowMeans(abCFup$surv)
# A plot to compare all three Absolute risk (Commulative Incidence)
plot(cbCIAve[,1],cbCIAve[,2],type='l',col="black",lwd=3,main="Cumulative Incidence comparison",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
lines(abCF,col="red",fun="event",lwd=3,conf.int = FALSE)
lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
lines(abCFup ,col="purple",fun="event",lwd=3,conf.int = FALSE)
legend("bottomright", 
       legend = c( "Penalized Weibull with casebase","Coxnet","Cox","Kaplan-Meier"), 
       col = c("black","red","Blue","purple"),
       lty = c(1, 1, 1), 
       bg = "gray90")
```

```{r covariatesRegularizedLollipopPlot,eval=FALSE}

#get covariates excluding d.time
estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
estimatescb$Model="casebase"
colnames(estimatescb)<-c("Coefs","Estimate","Model")
estimatescox=setDT(as.data.frame(coef(coxglmFit)[-c(1),1]), keep.rownames = TRUE)[]
estimatescox$Model="Cox"
colnames(estimatescox)<-c("Coefs","Estimate","Model")
estimates=rbind(estimatescb, estimatescox)
estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
library(ggstance)# for vertical dodge
  ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
        geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
        geom_point(position=ggstance::position_dodgev(height=0.2))

```

@@

Although the previous case studies provide a broad overview of the capabilities of \pkg{casebase}, they all have two properties in common:
\begin{itemize}
  \item All covariates are fixed, i.e. they do not change over time;
  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
\end{itemize}

Case-base sampling has been used in the literature to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case]. Hence, the main covariate of interest, i.e. exposure to the vaccine, was changing over time. In this context, case-base sampling offers an efficient alternative to nested case-control designs or self-matching.

The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above. To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{survival} package.

Recall the setting of this study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment. After enrollment, the program searches for a suitable donor for the patient, which can take anywhere between a few days to almost a year. We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occurred.

@@

In this article, we presented the \proglang{R} package \pkg{casebase} that provides functions to fit smooth parametric hazards and estimate cumulative incidence functions using case-base sampling. We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.

<!-- In the following table we provide a comparison between the Cox model and case-base sampling: -->

```{r, echo=FALSE, eval = FALSE, results='asis'}
knitr::kable(data.frame(`feature` = c("model type", "time", "cumulative incidence", "non-proportional hazards","model testing","competing risks", "prediction"),
           `Cox` = c("semi-parametric","left hand side of the equation", "step function", "interaction of covariates with time"," ", "difficult", "Kaplan-Meier-based"),
           `Case Base Sampling` = c("fully parametric (logistic/multinomial regression)", "right hand side - allows flexible modeling of time",
           "smooth-in-time curve", "interaction of covariates with time", "make use of GLM framework (LRT, AIC, BIC)",
           "cause-specific cumulative incidence functions (CIFs) directly obtained via multinomial regression", "ROC, AUC, risk reclassification probabilities")), escape = FALSE)
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)
options(kableExtra.latex.load_packages = FALSE)
library(tidyverse)
library(magrittr)
library(survival)
library(casebase)
library(splines)
library(tibble)
library(glmnet)
library(kableExtra)
library(pracma)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(1234)

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- TRUE
```

@@

In Hanley & Miettinen [-@hanley2009fitting], the authors suggest sampling the base series *uniformly* from the study base. In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base. More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism are proportional to the cardiovascular disease event rate given by the Framingham score. With this sampling mechanism, Saarela & Arjas are able to increase the efficiency of their estimators, when compared to uniform sampling.

<!-- The `casebase` package fits the family of hazard functions of the form -->

<!-- $$ h(t;X) = \exp[g(t;X)] $$ -->
<!-- where $t$ denotes time and $X$, the individual's covariate profile.  -->

Let $g(t; X)$ be the linear predictor such that $\log(h(t;X)) = g(t; X)$. Different functions of $t$ lead to different parametric hazard models. The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$:

$$ \log(h(t; X)) = \beta_0 + \beta_1 X.$$

The instantaneous failure rate is independent of $t$.\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the \textit{memoryless property} \citep{kalbfleisch2011statistical}.}

@@

At the time of writing, a cursory examination of CRAN's task view on survival analysis reveals that there are over 250 packages related to survival analysis [-@survTaskView]. For the purposes of this article, we restricted our description to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and cumulative incidence curve estimation. By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages. These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}. In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models. The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with a description of some of the functionalities they offer.

Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche]. Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: \pkg{CoxRidge} [-@perperoglou], \pkg{rstpm2} [-@clements_liu] and \pkg{survival} [-@survival-package]. On the other hand, \pkg{casebase} provides penalized estimation in the context of parametric hazards. To our knowledge, this is the only package to offer this functionality. \textcolor{red}{Is this true? Because it seems to be a logical conclusion from your paragraph.}

Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@sharabiani_mahani_2019], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival}. The types of models they allow vary for each package. For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of time using splines, which allows flexible modeling of the hazard function. Moreover, \pkg{flexsurv} has the ability to estimate both scale and shape parameters for a variety of parametric families (see Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear model of covariates (including time). Therefore, our package allows the user to model the effect of time using splines, and through interaction terms involving covariates and time, it also allows user to fit time-varying coefficient models. However, we do not explicitly model any shape parameter, unlike \pkg{flexsurv}.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks. The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression. \textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}. Could you incorporate some of their discussion, with proper citations? The biggest omission right now is \pkg{cmprsk}, which we actually use below.}

@@

\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\

@@

In the next case study, we show how case-base sampling can  be used in the context of a competing risk analysis. For illustrative purposes, we will use the same data that was used in Scrucca *et al* [-@scrucca2010regression]. The data was downloaded from the first author's website, and it is also available as part of the \pkg{casebase} package.

```{r bmtcrr-data, eval = eval_cs2, message = FALSE}
library(casebase)
data(bmtcrr)
```

@@

```{r bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.", eval = eval_cs2}
library(ggplot2)
risk_all <- dplyr::bind_rows(
    data.frame(Time = time_points, 
               Method = "Case-base", 
               Risk = risk_cb[,2], 
               Disease = "ALL",
               stringsAsFactors = FALSE),
    data.frame(Time = time_points, 
               Method = "Case-base", 
               Risk = risk_cb[,3], 
               Disease = "AML",
               stringsAsFactors = FALSE),
    data.frame(Time = time_points,
               Method = "Fine-Gray",
               Risk = risk_fg$P1[1,],
               Disease = "ALL",
               stringsAsFactors = FALSE),
    data.frame(Time = time_points,
               Method = "Fine-Gray",
               Risk = risk_fg$P1[2,],
               Disease = "AML",
               stringsAsFactors = FALSE)
    )

@@

For the third case study, we will look at regularized estimation for the hazard function. This can be performed using a combination of case-base sampling and regularized logistic regression. To demonstrate this capability, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support]. The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital. Imputation was conducted using predictive mean matching and using the \pkg{mice} package in \proglang{R} [@mice]. Before imputation, there was a total of 1000 individuals and 35 variables. After imputation, we have 474 individuals. \textcolor{red}{How come we lose individuals during imputation?} Of these, 67.5% experienced the event of interest. <!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}. The original data was obtained from the website of the Department of Biostatistics at Vanderbilt University. The imputed dataset is available as part of the \pkg{casebase} package.

@@

<!--Hospital death as a covariate was removed, as this is directly informative of death. The glmnet package on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.--> As before, the first step is to visualize the incidence density. The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}. \textcolor{red}{This population-time plot is strange. The red dots never seem to go above 200.}

```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
data(support)
support_popTime <- popTime(support,
                           time = "d.time", event = "death")
plot(support_popTime)
```

Before we proceed with the analysis, we first remove hospital death as a covariate, as this is directly informative of death and cannot be used as a predictor. Then we select one observation at random. We will estimate the cumulative incidence curve for this individual. The other observations will be used to fit the hazard function.

```{r supportSetUp, echo = TRUE, eval = eval_cs3}
support <- dplyr::select(support, -hospdead)

# Randomly select one observation
indiv <- sample(1:nrow(support), size = 1)
```

Our main objective is to estimate the cumulative incidence of death for a given covariate profile. Given the number of covariates available, we will do this using regularized estimation. As \pkg{casebase} makes use of the \pkg{glmnet} package, we will use the function \code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}. Specifically, we input a matrix \code{y}, which contains the time and event variables, and a matrix \code{x}, which contains all other variables we would like to include in the model. For the SUPPORT dataset, all factor variables were converted into dummy variables.

```{r support-design, eval = eval_cs3}
# Create design matrix without the time variable
support_matrix <- model.matrix(death ~ . - d.time, 
                               data = support)
# Remove intercept term
support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]

# Create test and train dataset
support_test <- support_matrix[indiv,,drop = FALSE]

support_train <- support_matrix[-indiv,]
y <- as.matrix(support[-indiv, c("death", "d.time")])
```

We are now ready to estimate the hazard function. We use lasso penalization; this is achieved by setting \code{alpha} to 1. Also, note that the argument \code{formula_time} can be used to specify the function form for time; in our example, we model time linearly.

```{r CBhazard, echo = TRUE, eval = eval_cs3}
support_cb <- fitSmoothHazard.fit(support_train, y, 
                                  formula_time = ~ d.time, 
                                  family = "glmnet", 
                                  event = "death", time = "d.time", 
                                  alpha = 1, ratio = 100)
```

Then, we take our model object and use the \code{absoluteRisk} function to compute our estimate of the cumulative incidence function. Since \code{glmnet} fitted the penalized logistic regression for a range of penalty terms, we need to specify which one we want to use for our prediction. We select \code{s = "lambda.1se"}, as recommended.

```{r CBabsolute, echo = TRUE, eval = eval_cs3}
time_points <- seq(1, max(support$d.time), by = 0.5)
support_risk <- absoluteRisk(support_cb,
                             time = time_points,
                             newdata = support_test,
                             s = "lambda.1se", method = "numerical")
```

We will compare this curve to a regularized version of Cox regression, and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we can fit the Cox regression model using regularized estimation.

```{r coxHaz, echo = TRUE, eval = eval_cs3}
# Create survival object for glmnet
surv_glmnet <- survival::Surv(time = y[, "d.time"], 
                              event = y[, "death"])

support_cox <- glmnet::cv.glmnet(x = support_train, 
                                 y = surv_glmnet,
                                 family = "cox", alpha = 1)
```

However, \pkg{glmnet} does not provide any functionality for estimating the baseline hazard, and the functionalities in the \pkg{survival} package do not work on the output from \code{cv.glmnet}. To circumvent this issue and get an estimate of the baseline hazard, we followed the following steps. First, we fitted a non-regularized Cox model using only the variables selected by \pkg{glmnet}; the baseline hazard will be estimated using this object. Then, we replaced the estimates of the regression coefficients by the estimates from \pkg{glmnet}. Finally, this modified \code{coxph} object is fed to the \code{survival::survfit} function to obtain the estimated survival function.

```{r coxAbsolute, echo = TRUE, eval = eval_cs3}
# Identifying selected covariates
nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
                                  s = "lambda.1se")
nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
# Creating a new dataset that only contains the covariates selected by glmnet
support_cox_data <- as.data.frame(cbind(y,
                                        support_train[, nonzero_covariates_cox$X1]))
# Fitting a cox model
support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
                          data = support_cox_data)
# Replace the estimate with those obtained by glmnet
support_cox$coefficients <- nonzero_coef_cox@x

# Fitting CI curve for cox+glmnet
support_cox_risk <- survival::survfit(support_cox,
                                      newdata = as.data.frame(support_test),
                                      time = time_points,
                                      type = "breslow")
```

We used the \pkg{survival} package to calculate the Kaplan-Meier estimate of the survival function.

```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
km <- survival::Surv(time = support$d.time, 
                     event = support$death)
abKm <- survival::survfit(km ~ 1,
                          type = 'kaplan-meier')
```

Now that our three cumulative incidence functions have been calculated, we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and \pkg{casebase} have a lower estimated cumulative incidence by the end of the study, when compared to Kaplan-Meier. \textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different. At the very least, I think we should compare the select covariates.}

```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue."}
plot(support_risk, type='l', col = "black", lwd = 2, 
     main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
     xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
lines(support_cox_risk, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
legend("bottom", 
       legend = c("casebase", "Cox+glmnet", "KM curve"), 
       col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
```

@@

In the previous case studies, we only considered covariates that were fixed at baseline. In this next case study, we will use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-varying covariates. As an example that already appeared in the literature, case-base sampling was to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case]. Hence, the main covariate of interest, i.e. exposure to the vaccine, was changing over time. In this context, case-base sampling offers an efficient alternative to nested case-control designs or self-matching.

Recall the setting of Stanford Heart Transplant study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment. After enrollment, the program searched for a suitable donor for the patient, which could take anywhere between a few days to almost a year. We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occurred.

@@

In this article, we presented the \proglang{R} package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating cumulative incidence functions using case-base sampling. We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through four case studies.

### 2020-02-07 analysis/paper/paper.tex

This report was generated on 2020-02-07 10:31:46 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> Local:    intro /home/turgeonm/Documents/git_repos/cbpaper
#> Head:     [d5e7d09] 2020-02-07: add discussion of flexible sampling
\end{CodeOutput}
\end{CodeChunk}

>>>>

This report was generated on 2020-02-07 11:07:26 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> Local:    intro /home/turgeonm/Documents/git_repos/cbpaper
#> Head:     [f1e32e8] 2020-02-07: add discussion of flexible sampling
\end{CodeOutput}
\end{CodeChunk}
